---
title: "EDA"
output: html_document
date: "`r Sys.Date()`"
params:
  dm: dm
  data_tbl: data_tbl
  year: year
editor_options: 
  chunk_output_type: console
---

```{r echo=FALSE}
year <- params$year
if (is.null(year)) {
  rlang::abort(
    c(
      "Cannot render without a year.",
      i = "Did you call `render_report('eda.Rmd', .year = 2019)`?"
    )
  )
}
```

```{r echo=FALSE}
dm <- params$dm
data_tbl <- params$data_tbl
```

```{r setup, include=FALSE}
library(tidyverse)
library(corrr)
pkgload::load_all()
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, fig.align = "center", message = FALSE, warning = FALSE)
```

The chosen year is `r params$year`.

```{r eval=eval_if_is_installed_strict(c("DiagrammeRsvg", "htmltools"))}
dm_svg <- DiagrammeRsvg::export_svg(dm::dm_draw(dm))
htmltools::HTML(dm_svg)
```

## Data

```{r}
tbl <- 
  data_tbl |> 
  compute_dx_gap(e_inc_num, c_newinc) |>
  mutate(is_hbc = as.character(is_hbc)) |> 
  select(-year, -country)
```

```{r eval=FALSE}
tbl |> 
  arrange(is_hbc, country_code) |> 
  gt::gt()
```

```{r}
tbl |> 
  count() |> 
  gt::gt(caption = "Total number of records")
```

```{r}
tbl |> 
  pivot_longer(-c(country_code, is_hbc), names_to = "variable_name") |> 
  left_join(dxgap_master_list, join_by(variable_name)) |> 
  select(variable_name, definition, dataset, value) |> 
  compute_completion_rate(
    id_vars = c("variable_name", "definition", "dataset"), 
    digits = 3
  ) |> 
  select(-var_name) |> 
  relocate(dataset, definition, variable_name) |> 
  arrange(dataset, variable_name) |> 
  gt::gt()
```


```{r eval=FALSE}
tbl |> 
  skimr::skim()
```


## Correlations

```{r}
corr_df_raw <- 
  tbl |> 
  select(-country_code) |> 
  correlate() 
```

```{r}
corr_all <- 
  compute_correlation(tbl, who_dx_gap) |> 
  mutate(is_hbc = "99") # all
```

```{r}
corr_by_type <- compute_correlation(tbl, who_dx_gap, by = "is_hbc")
```

```{r}
corr_all_df <- 
  corr_by_type |> 
  bind_rows(corr_all) |> 
  left_join(
    select(dxgap_master_list, variable_name, definition), 
    by = join_by(term == variable_name)
  ) |> 
  relocate(definition, term, starts_with("who_dx_gap"))
```

```{r fig.height=12, fig.width=12}
corr_names_vec_labels <- c(
     "0" = "who_dx_gap_non_hbc", 
     "99" = "who_dx_gap_all", 
     "1" = "who_dx_gap_hbc"
)
corr_names_vec_levels <- c(
    "who_dx_gap_non_hbc" = "0", 
    "who_dx_gap_all" = "99", 
    "who_dx_gap_hbc" = "1"
)
corr_all_df |> 
  # mutate(term = forcats::fct_reorder(term, corr, .na_rm = FALSE)) |> 
  mutate(term = forcats::fct_relevel(term, sort)) |> 
  mutate(
    corr_cat = forcats::as_factor(is_hbc),
    corr_cat = forcats::fct_relevel(corr_cat, corr_names_vec_levels)
  ) |> 
  ggplot(aes(term, who_dx_gap)) +
  geom_col() +
  coord_flip() +
  theme_minimal() +
  facet_wrap(vars(corr_cat), labeller = labeller(corr_cat = corr_names_vec_labels))
```


```{r}
corr_all_df |>
  pivot_wider(
    id_cols = c(term, definition), 
    names_from = is_hbc,
    values_from = who_dx_gap
  ) |> 
  rename(all_of(corr_names_vec_levels)) |> 
  arrange(term) |> 
  gt::gt()
```

```{r fig.height=15, fig.width=15}
corr_df_raw |> 
  plot_corr()
```


## Dx Gap

The Dx Gap is computed as follows:

$who\_dx\_gap = (e\_inc\_num - c\_newinc) / e\_inc\_num$

```{r fig.height=12, fig.width=12}
tbl |> 
  # mutate(country_code = forcats::fct_reorder(country_code, who_dx_gap, .na_rm = FALSE)) |> 
  mutate(country_code = forcats::fct_relevel(country_code, sort)) |> 
  ggplot(aes(country_code, who_dx_gap, fill = is_hbc)) +
  geom_col() +
  scale_y_continuous(labels = scales::label_percent(scale = 1)) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = rel(0.45))) +
  coord_flip() +
    labs(
    y = paste("Dx Gap in", params$year),
    caption = "Source: WHO"
  ) +
  facet_wrap(vars(is_hbc), scales = "free") 
```

```{r}
plot_density(
  tbl,
  who_dx_gap,
  is_hbc,
  params$year,
  "Dx Gap in",
  "Source: WHO"
) +
  scale_x_continuous(labels = scales::label_percent(scale = 1))
```


```{r long}
tbl <- 
  tbl |> 
  pivot_longer(
    cols = -c(is_hbc, country_code, who_dx_gap), 
    names_to = "variable_name"
  )
```

## Population total

```{r eval=eval_if(params$data_tbl, "pop_total")}
tbl |> 
  filter(variable_name == "pop_total") |> 
  plot_density(
    value,
    is_hbc,
    params$year,
    "Total population in",
    "Source: World Bank"
  )
```


## Population density

### Density distrbution

```{r eval=eval_if(params$data_tbl, "pop_density")}
tbl |>
  filter(variable_name == "pop_density") |>
  plot_density(
    value,
    is_hbc,
    params$year,
    "Population density (people per sq. km of land area) in",
    "Source: World Bank"
  )
```

### Dx Gap

```{r eval=eval_if(params$data_tbl, "pop_density")}
tbl |>
  filter(variable_name == "pop_density") |>
  pivot_wider(names_from = variable_name, values_from = value) |> 
  plot_scatter(
    pop_density, 
    who_dx_gap, 
    params$year, 
    is_hbc, 
    "Population density (people per sq. km of land area) in", 
    "Source: World Bank"
  )
```

## Population urban

### Density distribution

```{r eval=eval_if(params$data_tbl, "pop_urban_perc")}
tbl |>
  filter(variable_name == "pop_urban_perc") |>
  plot_density(
    value,
    is_hbc,
    params$year,
    "Urban population (% of total population) in",
    "Source: World Bank"
  ) +
    scale_x_continuous(labels = scales::label_percent(scale = 1)) 
```

### Dx Gap

```{r eval=eval_if(params$data_tbl, "pop_urban_perc")}
tbl |>
  filter(variable_name == "pop_urban_perc") |>
  pivot_wider(names_from = variable_name, values_from = value) |> 
  plot_scatter(
    pop_urban_perc,
    who_dx_gap,
    params$year,
    is_hbc,
    "Urban population (% of total population) in",
    "Source: World Bank"
  ) +
    scale_x_continuous(labels = scales::label_percent(scale = 1)) +
    coord_fixed()
```

## GDP

### Density distribution

```{r eval=eval_if(params$data_tbl, "gdp")}
tbl |> 
  filter(variable_name == "gdp") |>
  plot_density(
    value,
    is_hbc,
    params$year,
    "GDP (current US$) in",
    "Source: World Bank"
  ) 
```

### Dx Gap

```{r eval=eval_if(params$data_tbl, "gdp")}
tbl |>
  filter(variable_name == "gdp") |>
  pivot_wider(names_from = variable_name, values_from = value) |>
  plot_scatter(
    gdp,
    who_dx_gap,
    params$year,
    is_hbc,
    "GDP (current US$) in",
    "Source: World Bank"
  )
```

## Budget

```{r}
budget_master <- 
  dxgap_master_list |> 
  filter(dataset == "budget") |> 
  select(variable_name, definition)

budget_cols <- 
  budget_master |> 
  pull(variable_name)
```

```{r eval=eval_if(params$data_tbl, budget_cols)}
budget_long <- 
  tbl |>
  filter(variable_name %in% budget_cols) |> 
  left_join(budget_master, join_by(variable_name)) |> 
  mutate(definition = str_squish(str_remove(definition, "\\(.+\\)")))
```


### Density distribution

```{r eval=eval_if(params$data_tbl, budget_cols), fig.height=15, fig.width=12}
plot_density(
  budget_long,
  value,
  is_hbc,
  params$year,
  "Budget data in",
  "Source: Global Fund"
) +
  facet_wrap(vars(definition), ncol = 2, scales = "free") 
```

### Dx Gap

```{r eval=eval_if(params$data_tbl, budget_cols), fig.height=15, fig.width=12}
plot_scatter(
  budget_long,
  value,
  who_dx_gap,
  params$year,
  is_hbc,
  "Budget data in",
  "Source: WHO"
) +
  facet_wrap(vars(definition), ncol = 2, scales = "free") 
```

## Community

```{r}
comm_master <- 
  dxgap_master_list |> 
  filter(dataset == "community_engagement") |> 
  select(variable_name, definition)

comm_cols <- 
  comm_master |> 
  pull(variable_name)
```

```{r eval=eval_if(params$data_tbl, comm_cols)}
comm_long <- 
  tbl |>
  filter(variable_name %in% comm_cols) |> 
  left_join(comm_master, join_by(variable_name)) |> 
  mutate(definition = str_squish(str_remove(definition, "\\(.+\\)")))
```


### Density distribution

```{r eval=eval_if(params$data_tbl, comm_cols), fig.height=15, fig.width=12}
plot_density(
  comm_long,
  value,
  is_hbc,
  params$year,
  "Community data in",
  "Source: WHO"
) +
  facet_wrap(vars(definition), ncol = 2, scales = "free") 
```

### Dx Gap

```{r eval=eval_if(params$data_tbl, comm_cols), fig.height=15, fig.width=12}
plot_scatter(
  comm_long,
  value,
  who_dx_gap,
  params$year,
  is_hbc,
  "Community data in",
  "Source: WHO"
) +
  facet_wrap(vars(definition), ncol = 2, scales = "free") 
```

## Laboratories

```{r}
labs_master <- 
  dxgap_master_list |> 
  filter(dataset == "laboratories") |> 
  select(variable_name, definition)

labs_cols <- 
  labs_master |> 
  pull(variable_name)
```

```{r eval=eval_if(params$data_tbl, labs_cols)}
sites_long <- 
  tbl |> 
  filter(variable_name %in% labs_cols) |> 
  left_join(labs_master, by = join_by(variable_name)) 
```

### Density distribution

```{r eval=eval_if(params$data_tbl, labs_cols), fig.height=15, fig.width=12}
plot_density(
  sites_long,
  value,
  is_hbc,
  params$year,
  "Community data in",
  "Source: WHO"
) +
  facet_wrap(vars(definition), ncol = 2, scales = "free")
```

### Dx Gap

```{r eval=eval_if(params$data_tbl, labs_cols), fig.height=15, fig.width=12}
plot_scatter(
  sites_long,
  value,
  who_dx_gap,
  params$year, 
  is_hbc,
  "Labs and site data in",
  "Source: WHO"
) +
  facet_wrap(vars(definition), ncol = 2, scales = "free")
```

## Devices

### Density distribution

```{r eval=eval_if(params$data_tbl, "total_numb_device")}
tbl |> 
  filter(variable_name == "total_numb_device") |> 
  plot_density(
    value,
    is_hbc,
    params$year, 
    "TB diagnostic devices in",
    "Source: Global Fund"
)
```

### Dx Gap

```{r eval=eval_if(params$data_tbl, "total_numb_device")}
tbl |> 
  filter(variable_name == "total_numb_device") |>
  plot_scatter(
    value,
    who_dx_gap,
    params$year, 
    is_hbc,
    "TB diagnostic devices in",
    "Source: Global Fund"
  )
```

## Expenditures

```{r}
exp_master <- 
  dxgap_master_list |> 
  filter(dataset == "expenditure_and_utilisation") |> 
  select(variable_name, definition)

exp_cols <- 
  exp_master |> 
  pull(variable_name)
```

```{r eval=eval_if(params$data_tbl, exp_cols)}
exp_long <- 
  tbl |> 
  filter(variable_name %in% exp_cols) |> 
  left_join(exp_master, by = join_by(variable_name)) 
```

### Density distribution

```{r eval=eval_if(params$data_tbl, exp_cols), fig.height=15, fig.width=12}
plot_density(
  exp_long,
  value,
  is_hbc,
  params$year,
  "Community data in",
  "Source: WHO"
) +
  facet_wrap(vars(definition), ncol = 2, scales = "free")
```

### Dx Gap

```{r eval=eval_if(params$data_tbl, exp_cols), fig.height=15, fig.width=12}
plot_scatter(
  exp_long,
  value,
  who_dx_gap,
  params$year, 
  is_hbc,
  "Labs and site data in",
  "Source: WHO"
) +
  facet_wrap(vars(definition), ncol = 2, scales = "free")
```
