---
title: "TB Diagnostic Gap Report"
subtitle: "The validation and determinants of Finddx's diagnostic gap measure"
author: "cynkra"
date: 2023-07-12
toc: true
format: 
   html:
     code-tools: true
     code-fold: true
     df-print: paged
execute: 
  warning: false
  cache: true
editor:
  render-on-save: true
---

```{r}
library(readr)
library(dplyr)
library(corrr)
library(ggplot2)
library(tidyr)
library(broom)

pkgload::load_all(".")

finddx_ts <- read_tidy_tb(type = "time_series")
finddx_fixed <- read_tidy_tb(type = "fixed")
```

# Validating the diagnostic gap
In order to verify the validity of the "Diagnostic Gap" measure, also referred to as "Dx Gap", provided by Finddx, we computed an alternative diagnostic gap using [data](https://www.who.int/teams/global-tuberculosis-programme/data) from the World Health Organization (WHO).

## Defining the diagnostic gap
The TB diagnostic gap is the _"the proportion of the population with the
condition who are undiagnosed"_ [^1]. In other words, we can define the
diagnostic gap as the gap between the estimated number of people who fall ill
with TB each year and the number of people who are officially reported as
having been diagnosed. This means that the diagnostic gap represents people
who have TB but are not officially diagnosed, due to limitations or challenges
in the healthcare system.

[^1]: [The Lancet Commission on diagnostics: transforming access to diagnostics](https://doi.org/10.1016/S0140-6736(21)00673-5)

## Data Acquisition
The data required to calculate a new measure of diagnostic gap from the WHO can
be found in two separate data sets. The estimated total number of TB cases
expected each year can be found in the 'e_inc_num' variable in the 'WHO TB
burden estimates [>1Mb]' dataset:

```{r}
who_estimates <- read_csv(
  "https://extranet.who.int/tme/generateCSV.asp?ds=estimates"
)

who_estimates <- who_estimates |>
  select(
    country,
    year,
    tb_estimated_cases = e_inc_num
  )

who_estimates
```

The total number of actual reported cases by health systems can be found in the
'c_newinc' variable in the 'Case notifications [>2Mb]' dataset:

```{r}
who_notifications <- read_csv(
  "https://extranet.who.int/tme/generateCSV.asp?ds=notifications"
)

who_notifications <- who_notifications |>
  select(
    country,
    year,
    tb_notified_cases = c_newinc
  )

who_notifications
```

## Calculating the diagnostic gap
For each country and year in the WHO datasets, the diagnostic gap can be computed as follows:

- diagnostic gap = (estimated cases - notified cases) / estimated cases

```{r}
who_dx_gap <- who_estimates |>
  inner_join(who_notifications, by = c("year", "country")) |>
  mutate(
    who_dx_gap = (tb_estimated_cases - tb_notified_cases) / tb_estimated_cases
  ) |>
  select(country, year, who_dx_gap)

who_dx_gap
```

## Examining Finddx's diagnostic gap measure
Before we can use the newly calculated diagnostic gap measure from WHO to validate Finddx's gap measure, we must first analyse the distribution of Finddx's diagnostic gap measure.

```{r}
#| fig-cap: "Finddx's diagnostic gap measure demonstrates a normal distribtuion with the exception of an outlier at 1.25"
finddx_fixed |>
  ggplot(aes(rate_dx_gap)) +
  geom_density(fill = "steelblue", alpha = .5) +
  theme_minimal() +
  labs(x = "Finddx's diagnostic gap")
```

An outlier, with a diagnsoptic gap rate of 1.25, appears to be causing the data
to be positively skewed. Further inspection reveals this outlier comes from data
for the Russian Federation:

```{r}
finddx_fixed |>
  select(country, rate_dx_gap) |>
  arrange(desc(rate_dx_gap)) |>
  slice(1)
```

This value for the Russian Federation, with a diagnostic gap above 1, implies
that there were a negative number of TB cases notified. It is reasoned that this
is therefore a data entry or calculation error, and will be removed from the
analyses.

```{r}
finddx_fixed <- finddx_fixed |>
  filter(country != "Russian Federation")
```

::: {.callout-tip}
## Question for Finddx

Is the diagnostic gap value for the Russian Federation a mistake, or is there
an alternative reason for why it may be above 1?
:::


## Validating Finddx's diagnostic gap measure
Finddx's diagnostic gap measure can now be validated against the measure we
have just calculated from the WHO data. The Pearson's correlation coefficient
between the two measures of -0.986 indicates a very strong relationship,
validating the measure used by Finddx:

```{r}
who_dx_gap_2020 <- who_dx_gap |>
  filter(year == 2020) |>
  mutate(
    country = if_else(
      country == "Democratic Republic of the Congo",
      "DRC",
      country
    )
  )

dx_gaps_joined <- finddx_fixed |>
  select(country, finddx_dx_gap = rate_dx_gap) |>
  left_join(who_dx_gap_2020) |>
  select(
    country,
    `Finddx diagnostic gap` = finddx_dx_gap,
    `WHO diagnostic gap` = who_dx_gap
  )

dx_gaps_joined |>
  correlate() |>
  slice(1) |>
  select(-`Finddx diagnostic gap`)
```

This strong relationship can also be seen on the following scatter plot:

```{r}
#| fig-cap: Finddx's measure of diagnostic gap strongly negatively correlates with an alternative measure calculated from WHO data
dx_gaps_joined |>
  ggplot(aes(`Finddx diagnostic gap`, `WHO diagnostic gap`)) +
  geom_point(colour = "steelblue", size = 5, alpha = .5) +
  theme_minimal()
```

::: {.callout-tip}
## Question for Finddx

Why is the correlation between the WHO calculated measure and Finddx's measure
negative, implying the inverse of the diagnostic gap has been calculated?
:::

# Exploring the determinants of the diagnostic gap
An exploxatory analysis on the 'TB Policy' Excel dashboard supplied by Finddx
was conducted.

## Data selection
To intiate the explatory analysis, it was decided that a cross-sectional
analysis, looking at a single time point, would first be conducted. It was
decided that data for 2019 would be used for two reasons:

1. The data supplied by Finddx contains many missing values, especially for the
   labatory variables, for 2020.
2. The global effects of the Covid-19 pandemic began to take effect in
   2020 [^2] and the effects on the TB data presented in the supplied data set
   are unkown.

[^2]: [Timeline: WHO's COVID-19 response](https://www.who.int/emergencies/diseases/novel-coronavirus-2019/interactive-timeline#!)

The supplied data only included a measure of diagnositc gap for 2020, so the
value calculated from the 2019 WHO data is used from this point forward.

```{r}
who_dx_gap_2019 <- who_dx_gap |>
  filter(year == 2019) |>
  mutate(
    country = if_else(
      country == "Democratic Republic of the Congo",
      "DRC",
      country
    )
  )

finddx_2019 <- finddx_ts |>
  filter(year == 2019) |>
  pivot_wider(names_from = indicator, values_from = value) |>
  left_join(finddx_fixed) |>
  select(-year, -rate_dx_gap)

data_2019 <- finddx_2019 |>
  left_join(who_dx_gap_2019)
```

## Laboratory variables

### Distributions
First, the relationship between laboratory tests and diagnostic gap were
explored, as Finddx expressed this as a key area of interest.

```{r}
labs_2019 <- data_2019 |>
  select(country, population_100k, who_dx_gap, starts_with("labs_"))
```

The distribution of the laboratory test variables was inspected:

```{r}
#| fig-cap: "Distributions for different TB laboratory test counts"
labs_2019 |>
  select(-population_100k, -who_dx_gap) |>
  rename(
    `Labs performing smear tests` = labs_performing_smear_tests,
    `Labs performing xpert tests` = labs_performing_xpert_tests,
    `Labs performing culture liquid tests` = labs_performing_culture_tests_liquid,
    `Labs performing DST` = labs_performing_dst
  ) |>
  pivot_longer(!country) |>
  ggplot(aes(value)) +
  geom_density(fill = "steelblue", alpha = .5) +
  facet_wrap(vars(name), scales = "free") +
  theme_minimal()
```

A clear outlier with a value of 23,822 for laboratories performing culture
liquid tests can be observed in the distribtuion for labs performing culture 
liquid tests. This value comes from China and is ~10-fold higher than previous 
years:

```{r}
finddx_ts |>
  filter(
    country == "China",
    indicator == "labs_performing_culture_tests_liquid"
  )
```

The mean of previous years shall be used for this outlier from this point
forward.

```{r}
china_culture_tests_average <- finddx_ts |>
  filter(
    country == "China",
    indicator == "labs_performing_culture_tests_liquid"
  ) |>
  filter(year < 2019) |>
  summarise(mean = mean(value)) |>
  pull(mean)

labs_2019 <-
  labs_2019 |>
  mutate(
    labs_performing_culture_tests_liquid = if_else(
      country == "China",
      china_culture_tests_average,
      labs_performing_culture_tests_liquid
    )
  )
```

All of the laboratory variables demonstrated a positive skew. Given the
difference in population sizes between India and China and the rest of the
countries listed, this distribution is unsurprising. Indeed, a similar
distribution can be observed for population:

```{r}
#| fig-cap: "The distribution of populations demonstrates a positive skew"
labs_2019 |>
  ggplot(aes(population_100k)) +
  geom_density(fill = "steelblue", alpha = .5) +
  theme_minimal() +
  labs(x = "Population (100k)")
```

Given the diagnostic gap provides a relative measure (i.e., per capita),
the laboratory variables will now be normalised by population from this point
forwards.

```{r}
labs_2019_normalised <- labs_2019 |>
  mutate(
    across(
      starts_with("labs_"),
      ~ .x / population_100k,
      .names = "{.col}_per_100k"
    )
  ) |>
  select(country, who_dx_gap, ends_with("per_100k"))
```

The population normalised distributions can then be inspected:

```{r}
#| fig-cap: "The distribution of population normalised variables still demonstrate a positive skew"
labs_2019_normalised |>
  select(-who_dx_gap) |>
  rename(
    `Labs performing smear tests per 100k` = labs_performing_smear_tests_per_100k,
    `Labs performing xpert tests per 100k` = labs_performing_xpert_tests_per_100k,
    `Labs performing culture liquid tests per 100k` = labs_performing_culture_tests_liquid_per_100k,
    `Labs performing DST per 100k` = labs_performing_dst_per_100k
  ) |>
  pivot_longer(!country) |>
  ggplot(aes(value)) +
  geom_density(fill = "steelblue", alpha = .5) +
  facet_wrap(vars(name), scales = "free") +
  theme_minimal()
```

::: {.callout-note}
# Consideration for future analyses

Should the labatory variable distributions be transformed to a normal
distribubtion? (e.g., by applying a log transform)
:::

### Relationship to diagnostic gap
The correlation between the population normalised laboratory variables and the
calculated WHO diagnostic gap can be seen in table and plots below:

```{r}
labs_2019_normalised |>
  rename(
    `WHO Diagnostic Gap` = who_dx_gap,
    `Labs performing smear tests per 100k` = labs_performing_smear_tests_per_100k,
    `Labs performing xpert tests per 100k` = labs_performing_xpert_tests_per_100k,
    `Labs performing culture liquid tests per 100k` = labs_performing_culture_tests_liquid_per_100k,
    `Labs performing DST per 100k` = labs_performing_dst_per_100k
  ) |>
  correlate() |>
  select(term, `WHO Diagnostic Gap`) |>
  arrange(`WHO Diagnostic Gap`) |>
  filter(term != `WHO Diagnostic Gap`)
```

```{r}
#| fig-cap: "The correlation between laboratory tests and diagnostic gap varies"
labs_2019_normalised |>
  rename(
    `WHO Diagnostic Gap` = who_dx_gap,
    `Labs performing smear tests per 100k` = labs_performing_smear_tests_per_100k,
    `Labs performing xpert tests per 100k` = labs_performing_xpert_tests_per_100k,
    `Labs performing culture liquid tests per 100k` = labs_performing_culture_tests_liquid_per_100k,
    `Labs performing DST per 100k` = labs_performing_dst_per_100k
  ) |>
  pivot_longer(!c(country, `WHO Diagnostic Gap`)) |>
  ggplot(aes(value, `WHO Diagnostic Gap`)) +
  geom_point(colour = "steelblue", alpha = .8) +
  facet_wrap(vars(name), scales = "free") +
  theme_minimal()
```

```{r}
labs_2019_normalised_model <- lm(
  who_dx_gap ~ labs_performing_culture_tests_liquid_per_100k,
  data = labs_2019_normalised,
)

tidy(labs_2019_normalised_model)
```

Collectively, these data indicate that the number of labs performing culture
smear liquid tests might be a significant determinant of a countires diagnostic
gap, explaining up to 27.5% of the variance. The model suggests an inverse
relationship between the number of labs performing liquid culture tests per
100k population and the TB diagnostic gap.

::: {.callout-tip}
## Question for Finddx

Are you able to expand your data set beyond the listed 30 countries to improve
the strength of the models?
:::

## Other testing variables

### Distributions

Two other test variables, the rate of people tested bacteriologically, and the
rate of people tested with rapid diagnostics, can be found in the data supplied
from Finddx.

```{r}
other_tests_2019 <- data_2019 |>
  select(
    country,
    who_dx_gap,
    rate_bacteriologically_tested,
    rate_tested_with_rapid_dx
  ) |>
  mutate(
    rate_tested_with_rapid_dx = if_else(
      rate_tested_with_rapid_dx == "NR", # It is assumed that NR = "No record"
      NA_character_,
      rate_tested_with_rapid_dx
    ),
    rate_tested_with_rapid_dx = as.double(rate_tested_with_rapid_dx)
  )
```

The distribution of these other two test variables were inspected:

```{r}
#| fig-cap: "Distributions for other TB laboratory tests"
other_tests_2019 |>
  select(-who_dx_gap) |>
  rename(
    `Tested bacteriologically` = rate_bacteriologically_tested,
    `Tested with rapid diagnostic` = rate_tested_with_rapid_dx
  ) |>
  pivot_longer(!country) |>
  ggplot(aes(value)) +
  geom_density(fill = "steelblue", alpha = .5) +
  facet_wrap(vars(name), scales = "free") +
  theme_minimal() +
  labs(x = "Percentage (%)")
```

### Relationship to diagnostic gap
The correlations between the other laboratory variables and the calculated WHO
diagnostic gap can be seen in table and plots below:

```{r}
other_tests_2019 |>
  rename(
    `WHO Diagnostic Gap` = who_dx_gap,
    `Tested bacteriologically` = rate_bacteriologically_tested,
    `Tested with rapid diagnostic` = rate_tested_with_rapid_dx
  ) |>
  correlate() |>
  select(term, `WHO Diagnostic Gap`) |>
  arrange(desc(`WHO Diagnostic Gap`)) |>
  filter(term != `WHO Diagnostic Gap`)
```

```{r}
#| fig-cap: "The percentage of people bacteriologically tested  positively correlates with diagnostic gap, whereas those tested with rapid diagnositcs show no trend"
other_tests_2019 |>
  rename(
    `WHO Diagnostic Gap` = who_dx_gap,
    `Tested bacteriologically` = rate_bacteriologically_tested,
    `Tested with rapid diagnostic` = rate_tested_with_rapid_dx
  ) |>
  pivot_longer(!c(country, `WHO Diagnostic Gap`)) |>
  ggplot(aes(value, `WHO Diagnostic Gap`)) +
  geom_point(colour = "steelblue", alpha = .8) +
  facet_wrap(vars(name), scales = "free") +
  theme_minimal() +
  labs(x = "Percentage (%)")
```

```{r}
other_tests_2019_normalised_model <- lm(
  who_dx_gap ~ rate_bacteriologically_tested,
  data = other_tests_2019
)

tidy(other_tests_2019_normalised_model)
```

Neither variable appears to be a significant determinant of the diagnostic gap.